COVID-19 infection and attributable mortality in UK care homes: Cohort study using active surveillance and electronic records (March-June 2020)

### 2021-02-10 manuscript/manuscript.Rmd
last minute word cutting

Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 care homes for older people, which provide residential care with or without on-site nursing.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility. At the peak of the pandemic, deaths recorded in UK care home residents were three times higher than during the preceding year.[@onsDR2020] Staff in UK care homes also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK national statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested. Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on care homes.

Public health strategies to protect the public, residents and staff from SARS-CoV-2 include rapid testing, restriction of visitors, and vaccination. Such strategies require knowledge of the burden of and risk factors in residents and staff in care homes, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave. Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic in the context of limited testing capacity. There is no syndromic surveillance for infection in care homes in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in care homes until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per care home. Consequently national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in residents and staff in care homes.

To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in care homes to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections. We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest for-profit providers of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection. Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection. We also estimated mortality attributable to COVID-19.

@@

Staff and residents living/working in care homes for older people run by the Four Seasons Healthcare Group (FSHCG) between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")` were eligible for study inclusion. The FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is predominantly state-funded. Most residents are permanent, but a small proportion receive temporary (respite) care.

In 2020, there were 9,568 beds, representing 9% of all registered care home beds in England, Scotland and Northern Ireland (supplementary methods). 90% of FSHCG care homes participated in the whole care home testing programme, implying that all staff and residents were tested for COVID-19 at least once between 11 May and 22 June 2020.

@@

FSHCG collects electronic records on residents occupying 'private' beds, excluding those occupying beds that are 'block contracted' to the local authority (855 beds, see Figure 1). Records includes: dates of entry to and exit from the care home, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, older residents). Individual-level data on incidents including infections are reported via 'Datix': resident names, care home identifier, incident date/time, date of birth, sex, COVID-19 symptoms (9 multiple choices), test results, resident current location (care home/hospital), and death. Individual-level data on residents were linked to Datix reports (supplementary methods), and used to categorise residents' infection status into four groups: (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary methods). The term 'confirmed' denoted a positive PCR test. Datix was also used to differentiate deaths in hospital from those in the care home, and to identify COVID-19 related deaths. 1,492/1,880 (79%) of Datix reports were successfully linked.

@@

On 24 March FSHCG introduced a new reporting system requiring managers of each care home to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (new symptomatic cases, new confirmed cases). The number of occupied beds in each care home was reported weekly via the same mechanism. COVID-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to COVID-19 by the coroner. Data on staff deaths were not extracted due to the small number of cases.

Characteristics of each care home (number of beds, region, nursing versus residential care) were obtained from FSHCG.

@@

Risk factors considered included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (number of beds, occupancy, bed-to-staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]). Baseline care home occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable. We also estimated the ratio of beds to staff as a continuous variable.

To adjust for case under-ascertainment, we created a dummy variable indicating care homes that had experienced an outbreak, defined throughout the manuscript as care homes that recorded ≥1 confirmed infection or COVID-19 related death during the study period. This definition was preferred over a standard outbreak definition (≥2 cases linked in time and place) to compensate for poor COVID-19 case-ascertainment during the pandemic due to limited testing. Sensitivity analysis was used to assess the impact on excess mortality of using a more specific outbreak definition (supplementary methods).

@@

Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies. These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach, as described in supplementary methods. Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation. The total number of residents at risk of infection was unknown, so it was approximated in a multiple-decrement life table (supplementary methods). The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset. The incidence rate ratio for care home (based on aggregate data) versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]

Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (Table S1b). Due to this, individual-level data were only used to estimate age/sex-specific rates of infection and Cox proportional hazards models testing the association with individual and organisational-level risk factors.

@@

The aggregate dataset was used to estimate the crude rate of COVID-19 related mortality in residents and individual-level data were used to estimate rates of all-cause mortality and case-fatality by age and gender.

In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in 'non-outbreak' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.

A Cox proportional hazards regression model tested the effect of individual- and care home-level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.  We estimated the attributable fraction of deaths for each infection category in care homes with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak care homes. This fraction was obtained by using the model to predict the counterfactual mortality, then computing the attributable fraction within study.[@Samuelsen2008] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits. Huber sandwich estimators of variance accounted for the design effect of care home clustering in regression models.

Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. Computer scripts are available online.[@Dutey2020]

@@

The study included `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland and `r formatbm(desc_staff$total_staff)` staff. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one COVID-19 outbreak in either the individual-level or aggregate datasets.  The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset, and `r round(mean(residents_overview$resident_days))` days in the individual-level dataset.

@@

Care home managers recorded symptoms of infection in  `r formatbm(desc_residents$CT_total_symptomatic)` residents, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (Table S1a, Figure 2). An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital. The cumulative incidence of confirmed infection was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`]. The rate of confirmed infections in care homes in England was 13-fold higher in care homes compared to the community prevalence of infection derived from the ONS household COVID-19 infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]

@@

Incidence estimates derived from individual-level Datix records are reported in supplementary Table S1b.

@@

`r desc_residents$DAT_CFR$y` deaths occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate in infected residents (Groups C and D) of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] (Table S4). The case-fatality rate increased with age and was higher in men compared to women.

@@

Factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model. Male sex, age $\geq$ 85 years, and residence in a nursing care home (adjusted hazard ratio HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`). After controlling for organisational differences, care home size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for care homes with $\geq$ 70 beds versus <35 beds). Care home baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection. For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR = 1.3 [1.1; 1.5]).

@@

Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality (Figure S3). After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).

We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (group A) in non-outbreak care homes as the reference group. Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.2 [1.8; 2.6]). All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.8 [2.3; 6.4]) or symptomatic (Group D: adjusted HR = 14 [11; 18]). In confirmed infections, mortality was significantly higher in individuals with a record of symptoms.

@@

Model-based estimates indicate that 653/1,694 (39%) all-cause deaths were attributable to COVID-19 (Table S7). In care homes with outbreaks only, just 217/653 (33%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, Table S7).

@@

Across 179 care homes, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic. Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic. In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy may be important to reduce the spread of infection.

Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of care homes in England.[@onsVivaldi2020] However, both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.

Two thirds of care homes in our study reported at least one case of infection or death, in agreement with a recent study from the Lothian region of Scotland which reported that 61% of care homes for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system, and supports our assumption that residents in non-outbreak care homes had not been exposed to infection. This assumption made it possible to estimate mortality attributable to COVID-19.

Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of COVID-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.  Studies from other healthcare settings[@Lai2020; @Weinberger2020] have highlighted the ways in which COVID-19 has impacted delivery of care associated with excess mortality in individuals who are uninfected. Detailed analysis of cause of death and reasons for hospital admission in care home residents will be important to understand how the pandemic has affected the quality of care in care homes. Our analysis provides a method that could be widely applied to estimate excess mortality, provided care homes with outbreaks can be reliably identified.

In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and care home baseline occupancy. We also found staffing levels to be negatively associated with infection rates. These organisational factors, linked to chronic underfunding of the care sector, are likely to hinder the implementation of robust infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning. When staff care for fewer residents they also have reduced likelihood of spreading infection between residents. Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced care homes. These associations may also be confounded by other care home characteristics which could not be measured in this study, such as access to personal protective equipment, or the structure and layout of the building.

@@

The unique surveillance system we established in partnership with FSHCG tracked infections throughout the pandemic across a large number of care homes. To our knowledge, this is the most complete reporting system for COVID-19 infections in care homes published to date. It is possible that care homes that paid less attention to active surveillance to support control will have had higher levels of uncontrolled outbreaks compared to those seen in this study.

Our estimates of mortality attributable to COVID-19 are dependent on our definition of 'outbreak' versus 'non-outbreak' care homes. We used a sensitive definition of outbreaks (≥1 case per care home) because case under-ascertainment in care homes was common due to lack of testing.  However, it is possible that we incorrectly classified some care homes with one or a handful of cases throughout the pandemic as having experienced outbreaks. Other key limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses. We lacked information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity. The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each care home. In addition, our measures of care home occupancy were based on the pre-pandemic period and did not take account of higher vacancy rates during follow-up.

@@

UK numbers of infected residents and staff were underestimated during the first wave of the pandemic. Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in care homes to reduce occupancy and increase staffing. Although this presupposes increased use of testing to improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, because the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents. Since the incubation period and serial interval of COVID-19 is short,[@KaiWang2020] the interval between successive screens required to interrupt transmission may also need to be short. Strengthened surveillance in care homes would be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays,[@Flower2020] provided the predictive value of these tests is adequate.

>>>>

Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in care homes.[@PHE2020] In the UK, an estimated 400,000 residents live in approximately 11,000 care homes for older people, which provide residential care with or without on-site nursing.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces. At the peak of the pandemic, deaths recorded in UK care homes were three times higher than during the preceding year.[@onsDR2020] Staff also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK statistics suggests two-thirds of excess deaths recorded in residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested. Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on care homes.

Strategies to protect residents and staff from SARS-CoV-2 include rapid testing, restriction of visitors, and vaccination. These require knowledge of the burden of and risk factors for infection in residents and staff in care homes, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave. Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic due to limited testing capacity. There is no syndromic surveillance for infection in care homes in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's outbreak investigations which permitted a maximum of five tests per care home. Consequently, national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in care home residents and staff.

To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in care homes to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections. We analysed electronic health records from the Four Seasons Health Care Group (FSHCG), one of the UK’s largest for-profit providers of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection. Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection. We also estimated mortality attributable to COVID-19.

@@

Staff and residents living/working in care homes for older people run by the FSHCG between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")` were eligible for study inclusion. FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is predominantly state-funded. Most residents are permanent, but a small proportion receive temporary (respite) care.

In 2020, there were 9,568 beds, representing 9% of all registered care home beds in England, Scotland and Northern Ireland (supplementary methods). 90% of FSHCG homes participated in the whole care home testing programme, implying that all staff and residents were tested for COVID-19 at least once between 11 May and 22 June 2020.

@@

FSHCG collects electronic records on residents occupying 'private' beds, excluding those occupying beds that are 'block contracted' to the local authority (855 beds, see Figure 1). Records include: dates of entry to and exit from the care home, sex, date of birth, type of stay (residential/nursing) and care (general/dementia/older residents). Individual-level data on incidents including infections are reported via 'Datix': resident names, care home identifier, incident date/time, date of birth, sex, COVID-19 symptoms (9 multiple choices), test results, resident current location (care home/hospital), and death. Individual-level data on residents were linked to Datix reports (supplementary methods), and used to categorise residents' infection status into four groups: (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary methods). The term 'confirmed' denoted a positive PCR test. Datix was also used to differentiate deaths in hospital from those in the care home, and to identify COVID-19 related deaths. 1,492/1,880 (79%) of Datix reports were successfully linked.

@@

On 24 March FSHCG introduced a new reporting system requiring managers of each care home to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (new symptomatic cases, new confirmed cases). The number of occupied beds in each care home was reported weekly via the same mechanism. COVID-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to COVID-19 by the coroner.

@@

Risk factors included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (nursing/residential, number of beds, occupancy, bed-to-staff ratio, Index of Multiple Deprivation [@Abel2016]) obtained from FSHCG. Baseline care home occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable. We also estimated the ratio of beds to staff as a continuous variable.

A dummy variable marked the time from which an outbreak occurred, defined throughout the manuscript as a care home recording ≥1 confirmed infection or COVID-19 related death. This definition was preferred over a standard outbreak definition (≥2 cases linked in time/place) to compensate for poor COVID-19 case-ascertainment during the pandemic due to limited testing. Sensitivity analysis using a more specific outbreak definition can be found in supplementary methods.

@@

Incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies, the trusted source used for national reporting of cases in all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach (supplementary methods). The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset. The incidence rate ratio for care home (based on aggregate data) versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]

Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting. Due to this, individual-level data were only used to estimate age/sex-specific rates of infection and Cox proportional hazards models testing the association with individual and organisational-level risk factors.

@@

The aggregate dataset was used to estimate the crude rate of COVID-19 related mortality in residents. Individual-level data were used to estimate rates of all-cause mortality and case-fatality by age and gender.

To investigate COVID-19 excess mortality, we made the assumption that residents in 'non-outbreak' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 mortality.[@Burton2020]

A Cox proportional hazards regression model tested the effect of individual- and home-level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.  We estimated the attributable fraction of deaths for each infection category in care homes with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak care homes. This fraction was obtained by using the model to predict the counterfactual mortality, then computing the attributable fraction within study.[@Samuelsen2008] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits. Huber sandwich estimators of variance accounted for the design effect of care home clustering in regression models.

Data were analysed in `R`3.5.0, `epitool`[@Aragon2020] and `survival`.[@therneau2015]. Computer scripts are available online.[@Dutey2020]

@@

The study included `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland and `r formatbm(desc_staff$total_staff)` staff. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded an outbreak.  The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset, and `r round(mean(residents_overview$resident_days))` days in the individual-level dataset.

@@

Care home managers recorded symptoms of infection in  `r formatbm(desc_residents$CT_total_symptomatic)` residents, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (Table S1a, Figure 2). An additional `r desc_residents$CT_total_confirmed` residents had confirmed infections, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital. The cumulative incidence of confirmed infection was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`]. The rate of confirmed infections in care homes in England was 13-fold higher in care homes compared to the community prevalence of infection derived from the ONS household infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]

@@

@@

`r desc_residents$DAT_CFR$y` deaths occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate in infected residents (Groups C and D) of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%] (Table S4). The case-fatality rate increased with age and was higher in men.

@@

Factors affecting rates of confirmed infections were investigated in Cox Proportional Hazard models (`r table_nums("tab_CoxPH_confirmed", display = "cite")`). Male sex, age ≥85 years, and nursing care (adjusted hazard ratio HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed infection. After controlling for organisational differences, care home size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for care homes with ≥70 beds versus <35 beds). Care home baseline occupancy and staffing ratios had the greatest effect on residents’ risk of infection. For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR = 1.5 [1.1; 2.1]); a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR = 1.3 [1.1; 1.5]).

@@

Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examined the relationship between infection status (groups A–D) and mortality (Figure S3). After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and nursing care (adjusted HR = 1.3 [1.1; 1.6]).

We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (Group A) in non-outbreak care homes as the reference group. Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.2 [1.8; 2.6]). All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.8 [2.3; 6.4]) or symptomatic (Group D: adjusted HR = 14 [11; 18]). In confirmed infections, mortality was significantly higher in individuals with a record of symptoms.

@@

Model-based estimates indicate that 653/1,694 (39%) all-cause deaths were attributable to COVID-19 (Table S7). In care homes with outbreaks only, just 217/653 (33%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D).

@@

Across 179 care homes, 22% of residents experienced symptoms while 10% had laboratory-confirmed infections, with a case-fatality rate of 35.7% across the first wave of the pandemic. Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks. Only one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic. In addition to the need for active surveillance and increased testing capacity, higher staff-to-resident ratios and reduced occupancy may be important to reduce the spread of infection.

Our estimates are comparable to a large survey of managers of care homes in England.[@onsVivaldi2020] Both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.

Two-thirds of care homes in our study reported ≥1 infection or death, in agreement with a study from one region of Scotland which reported that 61% of care homes had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCG's active surveillance system, and supports our assumption that residents in non-outbreak care homes had not been exposed to infection. This assumption made it possible to estimate mortality attributable to COVID-19.

Our findings of excess deaths in those with no direct evidence of infection may be due to under-ascertainment, direct effects of COVID-19 control measures on delivery of care, and/or indirect effects due to additional disruption caused by the outbreak.[@Lai2020; @Weinberger2020] Detailed analysis of cause of death and reasons for hospital admission in care home residents will be important to understand how the pandemic has affected the quality of care in care homes. Our analysis provides a method that could be widely applied to estimate excess mortality, provided care homes with outbreaks can be reliably identified.

In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and care home baseline occupancy. We also found staffing levels to be negatively associated with infection rates. These organisational factors, linked to chronic underfunding of the care sector, are likely to hinder the implementation of robust infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning. When staff care for fewer residents they also have reduced likelihood of spreading infection between residents. Higher staff-to-resident ratios may also decrease reliance on agency staff working across multiple settings, and indicate better resourced care homes. These associations may also be confounded by other characteristics which could not be measured in this study, such as access to personal protective equipment, or the building structure/layout.

@@

The unique surveillance system we established in partnership with FSHCG tracked infections across a large number of care homes. To our knowledge, this is the most complete reporting system for COVID-19 infections in care homes published to date. It is possible that care homes that paid less attention to active surveillance to support control will have had higher levels of uncontrolled outbreaks compared to those seen in this study.

Our estimates of mortality attributable to COVID-19 are dependent on our definition of 'outbreak' versus 'non-outbreak' care homes. We used a sensitive definition (≥1 case/home) due to under-ascertainment caused by the lack of testing. However, it is possible that we incorrectly classified some care homes with only a few cases throughout the pandemic as having experienced outbreaks. Other key limitations relate to the completeness of individual-level data. We lacked information on comorbidity and ethnicity, shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity. The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each care home. Finally, our measures of care home occupancy were based on the pre-pandemic period and did not take account of higher vacancy rates during follow-up.

@@

UK numbers of infected residents and staff were underestimated during the first wave of the pandemic. Our findings support disease control strategies which integrate public health surveillance and rapid testing with investment in care homes to reduce occupancy and increase staffing. Although testing will improve case ascertainment, frequent testing in care home residents may not always be desirable if the risk of infection is low, because the testing procedure (nasopharyngeal swabs) is invasive and may distress vulnerable residents. Since the incubation period and serial interval of COVID-19 is short,[@KaiWang2020] the interval between successive screens required to interrupt transmission may also need to be short. Strengthened surveillance in care homes could be greatly facilitated by the availability of near patient testing platforms, such as lateral flow immunoassays,[@Flower2020] provided the predictive value of these tests is adequate.
